Mergeable Types

### 2017-02-27 paper/model.tex
Sec. 2

{\bf Lists}. List, as an abstract data type, supports three
operations: \C{insert x i l}, to insert an element \C{x} at position
\C{i} in the list \C{l}, \C{delete i l}, to delete the element at
\C{i}'th position, and \C{subst i x l}, to substitute the element at
\C{i}'th position with \C{x}. Lists of mergeable items are mergeable
while preserving the following properties:
\begin{itemize}
  \item Elements present in both the merging lists will be present in
  the final list. No element is deleted unless it is deleted in at
  least one list.
  \item An element deleted in one list will not reappear in the merged
  list.
  \item The order of elements is retained. That is, if $x$ occurs
  before $y$ in one of the lists, and if both elements are present in
  the merged list, then $x$ occurs before $y$ also in the merged list.
  \item If an element $y$ substitutes $x$ in one list, then $y$ also
  substitutes $x$ in the merged list (i.e., $x$ is absent and $y$ is
  present), unless $x$ is deleted or substituted in the other list. In
  the former case, deletion wins, and neither $x$ nor $y$ occur in the
  merged list. In the latter case, if $x$ is substituted by a
  different element $z$ in the other list, then the merged list
  substitutes $x$ by a merge of $y$ and $z$, defined as per the merge
  semantics of the list element type.
\end{itemize}
Thus list is an example of a polymorphic mergeable data type that
demonstrates how mergeability can be composed. The compositionality is
reified in the the signature of a mergeable list library is shown in
Fig.~\ref{fig:mlist-sig}; observe that the type of items in the list
(\C{A.t}) is also required to be mergeable (\C{A.merge}).

@@

A signficant advantage of mergeable types is their composability. For
instance, a mergeble list of counters can be obtained simply by
instantiating \C{MList} for \C{Counter} (i.e., \C{MList(Counter)}). A
non-trivial demonstration of the utility of such composability is
given below.

@@

{\bf Shopping List}. A shopping list is an

This example, however, implicitly makes certain assumptions
about the model and the underlying system, such as the existence of a
single least common ancestor (LCA) for any pair of versions, the
ability to access a previous version on any branch, and mergeability
of any two concurrent versions. Enforcing these guarantees in a fully
decentralized distributed setting requires addressing non-trivial
theoretical challenges described in the next section. Unlike the
canvas, the process of writing merge functions for data types with
non-trivial invariants could be quite involved. The following sections
describe a principled approach to deriving merge functions for any
data type given its invariants. Like \C{Canvas.merge}, merge functions
often need to compare multiple versions of a data type for structural
equality, which, when performed naively, incurs cost linear in the
size of versions. The representation of a versioned data type in \name
is optimized for performing comparisons that allows structural
equality to be determined in constant time. Such engineering details,
which contribute to the practicality of our programming model are also
discussed in subsequent sections.

>>>>

{\bf Lists}. List, as an abstract data type, supports three
operations: \C{insert x i l}, to insert an element \C{x} at position
\C{i} in the list \C{l}, \C{delete i l}, to delete the element at
\C{i}'th position, and \C{subst i x l}, to substitute the element at
\C{i}'th position with \C{x}. Lists of mergeable items are mergeable
while preserving the following properties:
\begin{itemize}
  \item Elements present in both the merging lists will be present in
  the final list. No element is deleted unless it is deleted in at
  least one list.
  \item An element deleted in one list will not reappear in the merged
  list (unlike Amazon's replicated shopping cart~\cite{dynamo}).
  \item The order of elements is retained. That is, if $x$ occurs
  before $y$ in one of the lists, and if both elements are present in
  the merged list, then $x$ occurs before $y$ also in the merged list.
  \item If an element $y$ substitutes $x$ in one list, then $y$ also
  substitutes $x$ in the merged list (i.e., $x$ is absent and $y$ is
  present), unless $x$ is deleted or substituted in the other list. In
  the former case, deletion wins, and neither $x$ nor $y$ occur in the
  merged list. In the latter case, if $x$ is substituted by a
  different element $z$ in the other list, then the merged list
  substitutes $x$ by a merge of $y$ and $z$, defined as per the merge
  semantics of the list element type.
\end{itemize}
Thus list is an example of a polymorphic mergeable data type that
demonstrates how mergeability can be composed. The compositionality is
reified in the the signature of a mergeable list library is shown in
Fig.~\ref{fig:mlist-sig}; observe that the type of items in the list
(\C{A.t}) is also required to be mergeable (\C{A.merge}).

@@

\GK{Somewhere here, I want to say:} To better appreciate the virtues
of mergeable types, we encourage the reader to think about how one
might implement the mergeable list type with the explicit effect
representation described in Sec.~\ref{sec:intro}.

\begin{figure}
  \centering
  \includegraphics[scale=0.4]{Figures/shopping-list}

  \caption{Merging concurrent shopping lists}
  \label{fig:shopping-list}
\end{figure}

A significant advantage of mergeable types is their composability. As
demonstrated by \C{MList}, the merge semantics of a composed mergeable
type is a composition of merge semantics of the individual types.  For
instance, a mergeable list of integers can be obtained by
instantiating \C{MList} for an \C{Int} type, where \C{Int.merge} is
same as \C{Counter.merge} (Sec.~\ref{sec:intro}). 
An alternative instantiation of \C{MList} is for a quantity type
(\C{Qty}) that, like \C{Int}, supports \C{add} and \C{subtract}
operations, but defines its \C{merge} semantics a little differently:
\begin{ocaml}
let merge lca v1 v2 = if v1 = lca then v2 else (if v2 = lca then v1 else max v1 v2)
\end{ocaml}
\C{MList(Qty)}, like \C{MList(Int)}, is a list of integers, and has
the same semantics under a sequential execution. Their differences
however manifest under a concurrent execution, where both adopt
different methods to reconcile concurrent updates.  is given
below.

@@

{\bf Shopping List}. A non-trivial demonstration of composability of
mergeable types is a shopping list application. The application
allows its users to collaboratively build a grocery shopping
list (listing items in decreasing ), by adding new items with a
specified quantity, deleting existing items, and increasing or
decreasing their quantity. An item (\C{Item.t}) is represented as a
tuple of its name (a string) and the quantity (A \C{Qty.t}). Merge
function for \C{Item.t} merges the quantities of items (using
\C{Qty.merge}) with same names. Items with different names are
considered distinct. A shopping list is a mergeable list of items
(i.e., \C{MList(Item)}). \C{MList(Item).merge} automatically lifts the
merge semantics for items to merge semantics for lists of items.
Fig.~\ref{fig:shopping-list} illustrates. Here, two users, Alice and
Bob, concurrently edit a shopping list whose initial contents are
as milk and eggs. While both simultaneously update the quantity of
eggs, Bob also removes milk and inserts candy. The merged shopping
list contains eggs and candy, where the quantity of eggs is obtained
by merging Alice's and Bob's quantity as per \C{Qty.merge} semantics.

The examples described above make certain assumptions about the model
and the underlying system, such as the existence of a single least
common ancestor (LCA) for any pair of versions, the ability to access
a previous version on any branch, and mergeability of any two
concurrent versions. Enforcing these guarantees in a fully
decentralized distributed setting requires addressing non-trivial
theoretical challenges. These, and other details that contribute to
the practicality of our model, such as containing the complexity of
merges, are discussed in subsequent sections.
